---
title: "metadid Simulation Validation Report"
author: "metadid-sims pipeline"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
    page-layout: full
execute:
  echo: true
  warning: false

knitr:
  opts_chunk:
    fig.width: 16
    fig.height: 7
    out.width: "100%"
---

```{r setup}
library(targets)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)

tar_source("R/scenarios.R")
tar_source("R/plots.R")

all_agg <- tar_read(all_agg)
all_rep <- tar_read(all_rep)
scenario_lookup_tbl <- scenario_lookup()
```

## Executive Summary

```{r exec-summary}
# Flag scenarios with empirical coverage below 80% for the treatment effect
poor_coverage <- all_agg |>
  filter(parameter == "treatment_effect_mean", empirical_coverage < 0.80)

convergence_issues <- all_agg |>
  filter(parameter == "treatment_effect_mean", pct_rhat_bad > 0.10)
```

**`r nrow(all_agg |> distinct(scenario_id))`** scenarios evaluated across
**`r max(all_agg$n_reps, na.rm = TRUE)`** replications each.

`r if (nrow(poor_coverage) == 0) "All scenarios achieved ≥80% empirical coverage for the treatment effect mean." else paste0(nrow(poor_coverage), " scenario(s) had coverage below 80% for the treatment effect mean.")`

`r if (nrow(convergence_issues) == 0) "No convergence issues detected (all scenarios had <10% of fits with Rhat > 1.05)." else paste0(nrow(convergence_issues), " scenario(s) had >10% of fits with convergence warnings.")`

```{r poor-coverage-table}
#| eval: !expr nrow(poor_coverage) > 0
kable(
  poor_coverage |>
    left_join(
      tibble(
        scenario_id = names(SCENARIO_CONFIGS),
        description = purrr::map_chr(SCENARIO_CONFIGS, "description")
      ),
      by = "scenario_id"
    ) |>
    select(scenario_id, description, model_label, empirical_coverage, mean_bias, rmse),
  digits = 3,
  caption = "Scenarios with <80% coverage for treatment_effect_mean"
)
```

## A. Calibration Studies

These scenarios test the model's core calibration properties — coverage, bias, and RMSE — under a range of study counts, design mixes, and parameterisations. All use the full `meta_did` model unless noted. The baseline scenario (A1) uses 20 DiD studies with default parameters; subsequent scenarios probe how performance changes with sample size (A2–A4), mixed designs (A5–A6), covariates (A7), correlated effects (A8), unnormalised fitting (A9–A10), individual-level data (A11–A12), rho estimation strategies (A13), and sign/direction variants (A14–A18).

```{r a-scenarios}
scenario_summary_table("A") |>
  kable(caption = "Category A: scenario definitions")
```

```{r a-diagnostics}
#| fig-height: 9
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "A")
```

```{r a-table}
results_table(all_agg, scenario_lookup_tbl, "A")
```

## F. Large-N Bias Probes

These scenarios use large numbers of studies (200 total) to narrow posteriors and expose any systematic bias that would be obscured in smaller samples. They vary the design mix, normalisation strategy, and model specification, allowing a stringent test of whether the model is asymptotically unbiased.

```{r f-scenarios}
scenario_summary_table("F") |>
  kable(caption = "Category F: scenario definitions")
```

```{r f-diagnostics}
#| fig-height: 7
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "F")
```

```{r f-table}
results_table(all_agg, scenario_lookup_tbl, "F")
```

## B. Comparative Studies

These scenarios pit two model configurations against each other on the same simulated data, asking which recovers the treatment effect better under specific conditions. Each scenario has a clear hypothesis about which model should win (or whether they should tie).

```{r b-scenarios}
scenario_summary_table("B") |>
  kable(caption = "Category B: scenario definitions")
```

```{r b-diagnostics}
#| fig-height: 7
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "B")
```

```{r b-table}
results_table(all_agg, scenario_lookup_tbl, "B")
```

## C. Outlier and Heavy-Tailed Studies

These scenarios test robustness to departures from Gaussian study-level heterogeneity. The contaminating distributions are chosen to be detectable but not catastrophic, so that a robust model (Student-t heterogeneity) should outperform the normal model without being trivially obvious.

```{r c-scenarios}
scenario_summary_table("C") |>
  kable(caption = "Category C: scenario definitions")
```

```{r c-diagnostics}
#| fig-height: 8
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "C")
```

```{r c-table}
results_table(all_agg, scenario_lookup_tbl, "C")
```

## D. Assumption Violation Studies

These scenarios introduce deliberate mismatches between the DGP and the fitted model to quantify the cost of specific assumption violations — unmodelled design effects, heterogeneous within-study variance, misspecified rho, effect-size/sample-size correlation, and ignored RCT baseline imbalance.

```{r d-scenarios}
scenario_summary_table("D") |>
  kable(caption = "Category D: scenario definitions")
```

```{r d-diagnostics}
#| fig-height: 6
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "D")
```

```{r d-table}
results_table(all_agg, scenario_lookup_tbl, "D")
```

## E. Edge Cases

These scenarios probe boundary conditions: extreme pre-post correlations, zero heterogeneity, very large heterogeneity, unbalanced arms, single studies per design, and missing rho values. They test whether the model remains stable and well-calibrated when inputs approach the limits of the assumed data-generating regime.

```{r e-scenarios}
scenario_summary_table("E") |>
  kable(caption = "Category E: scenario definitions")
```

```{r e-diagnostics}
#| fig-height: 6
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "E", exclude = "E5")
```

```{r e-table}
results_table(all_agg, scenario_lookup_tbl, "E")
```

## Convergence Diagnostics

```{r convergence}
plot_convergence(all_agg)
```

## Full Results Table

```{r full-table}
all_agg |>
  arrange(scenario_id, model_label, parameter) |>
  kable(digits = 3, caption = "All aggregated results")
```
