---
title: "metadid Simulation Validation Report"
author: "metadid-sims pipeline"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
    page-layout: full
execute:
  echo: true
  warning: false

knitr:
  opts_chunk:
    fig.width: 16
    fig.height: 7
    out.width: "100%"
---

```{r setup}
library(targets)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)

tar_source("R/scenarios.R")
tar_source("R/plots.R")

all_agg <- tar_read(all_agg)
all_rep <- tar_read(all_rep)
scenario_lookup_tbl <- scenario_lookup()
```

## Executive Summary

```{r exec-summary}
# Flag deviations on the primary estimand (treatment_effect_mean), then split
# them into GENUINE issues vs deviations that are EXPECTED by design
# (misspecification demos, boundary/degenerate setups, normalisation stress
# tests). See scenario_expectations() in R/scenarios.R for the registry.
mean_agg <- all_agg |> filter(parameter == "treatment_effect_mean")

flagged <- bind_rows(
  mean_agg |> filter(empirical_coverage < 0.80) |>
    mutate(issue = "coverage < 80%"),
  mean_agg |> filter(pct_rhat_bad > 0.10) |>
    mutate(issue = ">10% of fits Rhat > 1.05")
) |>
  tag_expectations() |>
  left_join(scenario_lookup(), by = "scenario_id")

genuine  <- flagged |> filter(!expected)
expected <- flagged |> filter(expected)
```

**`r nrow(all_agg |> distinct(scenario_id))`** scenarios evaluated across
**`r max(all_agg$n_reps, na.rm = TRUE)`** replications each.

`r if (nrow(genuine) == 0) "**No genuine issues:** every deviation on the treatment-effect mean is expected by design (see below)." else paste0("**", nrow(genuine), " genuine issue(s)** on the treatment-effect mean need attention (below). A further ", nrow(expected), " flagged result(s) are expected by design.")`

### Genuine issues

```{r genuine-issues-table}
#| eval: !expr nrow(genuine) > 0
kable(
  genuine |>
    select(scenario_id, description, model_label, issue,
           empirical_coverage, mean_bias, rmse, pct_rhat_bad),
  digits = 3,
  caption = "Genuine issues on treatment_effect_mean (coverage < 80% or convergence)"
)
```

`r if (nrow(genuine) == 0) "_None._" else ""`

### Expected by design (not defects)

These deviations are intended — misspecification demonstrations, boundary/degenerate
setups, or deliberate normalisation stress tests — and are listed for completeness.

```{r expected-issues-table}
#| eval: !expr nrow(expected) > 0
kable(
  expected |>
    select(scenario_id, description, model_label, issue,
           empirical_coverage, mean_bias, reason),
  digits = 3,
  caption = "Expected-by-design deviations on treatment_effect_mean"
)
```

`r if (nrow(expected) == 0) "_None._" else ""`

## Paper Figure

The publication figures are **not** rendered in this report. Figure 1 and
Figure 2 are built by the `paper-figure` workflow, which simulates the
figure-only categories (K, M, Q, R, S, T) itself and publishes
`figure1`/`figure2` (PNG and PDF) plus `figure_panels.csv`,
`illustration_draws.csv` and `paired_rmse_MT.csv` as run artifacts.

Those categories are deliberately excluded from this nightly pipeline: they
are generated programmatically, so the scenario matrix in `sims.yaml` cannot
see them, and pulling them into `all_agg` would make this report's job
re-simulate every one of them serially on a single runner.

The nightly sweep aggregates are still exported as `output/figure_panels.csv`
(target `figure_data_csv`), tagged by scenario category.

## A. Calibration Studies

These scenarios test the model's core calibration properties — coverage, bias, and RMSE — under a range of study counts, design mixes, and parameterisations. All use the full `meta_did` model unless noted. The baseline scenario (A1) uses 20 DiD studies with default parameters; subsequent scenarios probe how performance changes with sample size (A2–A4), mixed designs (A5–A6), covariates (A7), correlated effects (A8), unnormalised fitting (A9–A10), individual-level data (A11–A12), rho estimation strategies (A13), and sign/direction variants (A14–A18).

```{r a-scenarios}
scenario_summary_table("A") |>
  kable(caption = "Category A: scenario definitions")
```

```{r a-diagnostics}
#| fig-height: 9
#| column: screen
#| out-width: "100%"
A_combined    <- tar_read(A_combined)
A_agg_combined <- tar_read(A_agg_combined)
plot_diagnostics(A_combined, A_agg_combined, scenario_lookup_tbl, "A")
```

```{r a-table}
results_table(all_agg, scenario_lookup_tbl, "A")
```

## F. Large-N Bias Probes

These scenarios use large numbers of studies (200 total) to narrow posteriors and expose any systematic bias that would be obscured in smaller samples. They vary the design mix, normalisation strategy, and model specification, allowing a stringent test of whether the model is asymptotically unbiased.

```{r f-scenarios}
scenario_summary_table("F") |>
  kable(caption = "Category F: scenario definitions")
```

```{r f-diagnostics}
#| fig-height: 7
#| column: screen
#| out-width: "100%"
F_combined     <- tar_read(F_combined)
F_agg_combined <- tar_read(F_agg_combined)
plot_diagnostics(F_combined, F_agg_combined, scenario_lookup_tbl, "F")
```

```{r f-table}
results_table(all_agg, scenario_lookup_tbl, "F")
```

## B. Comparative Studies

These scenarios pit two model configurations against each other on the same simulated data, asking which recovers the treatment effect better under specific conditions. Each scenario has a clear hypothesis about which model should win (or whether they should tie).

```{r b-scenarios}
scenario_summary_table("B") |>
  kable(caption = "Category B: scenario definitions")
```

```{r b-diagnostics}
#| fig-height: 7
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "B")
```

```{r b-table}
results_table(all_agg, scenario_lookup_tbl, "B")
```

## C. Outlier and Heavy-Tailed Studies

These scenarios test robustness to departures from Gaussian study-level heterogeneity. The contaminating distributions are chosen to be detectable but not catastrophic, so that a robust model (Student-t heterogeneity) should outperform the normal model without being trivially obvious.

```{r c-scenarios}
scenario_summary_table("C") |>
  kable(caption = "Category C: scenario definitions")
```

```{r c-diagnostics}
#| fig-height: 8
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "C")
```

```{r c-table}
results_table(all_agg, scenario_lookup_tbl, "C")
```

## D. Assumption Violation Studies

These scenarios introduce deliberate mismatches between the DGP and the fitted model to quantify the cost of specific assumption violations — unmodelled design effects, heterogeneous within-study variance, misspecified rho, effect-size/sample-size correlation, and ignored RCT baseline imbalance.

```{r d-scenarios}
scenario_summary_table("D") |>
  kable(caption = "Category D: scenario definitions")
```

```{r d-diagnostics}
#| fig-height: 6
#| column: screen
#| out-width: "100%"
D_combined     <- tar_read(D_combined)
D_agg_combined <- tar_read(D_agg_combined)
plot_diagnostics(D_combined, D_agg_combined, scenario_lookup_tbl, "D")
```

```{r d-table}
results_table(all_agg, scenario_lookup_tbl, "D")
```

## E. Edge Cases

These scenarios probe boundary conditions: extreme pre-post correlations, zero heterogeneity, very large heterogeneity, unbalanced arms, single studies per design, and missing rho values. They test whether the model remains stable and well-calibrated when inputs approach the limits of the assumed data-generating regime.

```{r e-scenarios}
scenario_summary_table("E") |>
  kable(caption = "Category E: scenario definitions")
```

```{r e-diagnostics}
#| fig-height: 6
#| column: screen
#| out-width: "100%"
E_combined     <- tar_read(E_combined)
E_agg_combined <- tar_read(E_agg_combined)
plot_diagnostics(E_combined, E_agg_combined, scenario_lookup_tbl, "E", exclude = "E5")
```

```{r e-table}
results_table(all_agg, scenario_lookup_tbl, "E")
```

## H. Time Trend Distributional Misspecification

The full model assumes study-level time trends are exchangeable draws from a normal distribution. These scenarios test what happens when that assumption is violated at the population level, using a 15 DiD + 15 PP design so that PP studies — which rely on trend estimation for effect identification — are directly exposed to the misspecification. Each scenario compares three models on the same simulated data: the **full normal** model (`meta_did`), the **full robust** model (`meta_did` with Student-t heterogeneity), and the **naive** model (`meta_did_general` with `time_trend = "fixed_zero"`).

H1 generates trends from a negated log-normal (mean = −0.04, log-SD = 1), producing strong right-skewness: most studies have trends close to zero with occasional large negative trends. H2 uses a 50/50 mixture of two normal components centred at −0.01 and −0.10 — a bimodal distribution the normal model cannot represent regardless of sample size. H3 keeps the normal DGP but inflates `sigma_trend` to 0.10 (five times the default), testing the regime where trend uncertainty dominates effect uncertainty. H4 introduces a design-confounded trend: DiD studies are small (`n ~ U(30, 70)`) and have more negative trends, while PP studies are large (`n ~ U(150, 250)`) and have trends closer to zero; because the model pools trends across designs, it over-adjusts PP outcomes.

```{r h-scenarios}
scenario_summary_table("H") |>
  kable(caption = "Category H: scenario definitions")
```

```{r h-diagnostics}
#| fig-height: 7
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "H")
```

```{r h-table}
results_table(all_agg, scenario_lookup_tbl, "H")
```

## G. Bias Source Investigation

Category G probes the two systematic bias sources identified in the large-N results (Category F). It comprises two sub-investigations.

**G1–G5: Prior sensitivity (unnormalised model).** The unnormalised model showed a small positive bias (~0.6–0.9%) for a negative true effect, consistent with hierarchical shrinkage toward zero from the default `N(0, 10)` prior on `treatment_effect_mean`. G1–G5 vary the prior standard deviation from 1 to 100 (G3 replicates the default) to test whether the bias decreases monotonically as the prior becomes more diffuse.

**G6–G9: Jensen's inequality (normalised model).** The normalised model showed ~1% systematic downward bias at 200 studies, consistent with Jensen's inequality on the ratio estimator: because the model fits E[θ/baseline] rather than E[θ]/E[baseline], the convexity of the reciprocal function induces a negative shift proportional to between-study baseline variance. G6–G8 vary between-study baseline variation (`baseline_sd` = 0, 0.01, 0.10) to confirm that the bias tracks the Jensen term; G9 uses a very large control group (`n_control = 5000`) to test whether high within-study precision alone eliminates the bias.

```{r g-scenarios}
scenario_summary_table("G") |>
  kable(caption = "Category G: scenario definitions")
```

```{r g-diagnostics}
#| fig-height: 9
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "G")
```

```{r g-table}
results_table(all_agg, scenario_lookup_tbl, "G")
```

## I. Multiplicative Covariate Scenarios

Category I tests the multiplicative-covariate feature on the metadid `multiplicative-covars` branch: an estimated per-level factor (`effect_multiplier[<level>]`, with the reference level fixed at 1) applied to the population-mean linear predictor `treatment_effect_mean + X_cov · beta_cov`. Per-study heterogeneity is centred on the multiplied mean. The feature is useful when some study settings represent an attenuated or amplified version of the underlying effect (e.g. experimental vs real-world conditions, drug formulation, study source).

Every I scenario pairs a modelled-mode fit (`normalise = TRUE`) with a raw-scale fit (`normalise = FALSE`) via the `compare` block so we can read the multiplier-recovery signal on both scales — the one exception is I7 (individual-level data), which runs the modelled arm only: on the raw scale the individual-data + multiplicative posterior fails to converge and hits max treedepth every iteration, so the arm was dropped (the raw multiplier signal is already covered by I1/I2/I8/I9). I5 and I6 additionally pair with-multiplier vs without-multiplier fits to quantify the misspecification cost of omitting the feature when truth has it (I5) and the false-positive cost of including it spuriously when truth doesn't (I6). All scenarios use a bespoke simulator (`simulate_multiplicative_levels`) that draws per-study true effects from `θ_i ~ N(m_{level(i)} · (μ_θ + X_cov_i · β_cov), σ_θ²)`.

I8 and I9 exercise the two-covariate product extension (`multiplicative_covariate = ~ level + level2`): each study's overall factor is the product of two independently estimated per-covariate factors, labelled `effect_multiplier[<column>:<level>]` in the posterior. I8 crosses two binary covariates — one attenuating (0.6), one amplifying (1.4), so the amplifying direction is covered too — and I9 crosses a binary attenuator with a three-level covariate (0.7, 0.4), testing simultaneous recovery of three factors from a fully crossed design. The key diagnostics are per-factor coverage and the absence of leakage between the two factors (each multiplier should recover its own truth, not a blend of the product).

```{r i-scenarios}
scenario_summary_table("I") |>
  kable(caption = "Category I: scenario definitions")
```

```{r i-diagnostics}
#| fig-height: 7
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "I")
```

```{r i-table}
results_table(all_agg, scenario_lookup_tbl, "I")
```

## X. Randomisation and Baseline Imbalance

The baseline difference $\gamma_i$ is the one nuisance parameter a post-only study cannot identify at all: its estimated effect is $\tilde\theta_i = \phi_i(1 + \tilde\beta_i) - \gamma_i$, so whatever the model assumes about imbalance subtracts directly from that study's treatment effect. These scenarios ask **whether imbalance learned from one kind of study should transport to another**.

This is a different question from category P, which varies the *size* of imbalance among post-only studies that are all unrandomised. Category X varies *which studies are randomised*.

Four fit arms recur throughout:

| Arm | `baseline_imbalance` | `mu_gamma` | Meaning |
|---|---|---|---|
| `shared` | `estimated` | `estimated` | One $\gamma$ population with an estimated mean, applied to every study. metadid's default before this work. |
| `zero_mean` | `estimated` | `zero` | One population, mean pinned at 0. Magnitude transports, direction does not. |
| `two_pop` | `by_randomisation` | `zero` | Non-randomised studies pooled; randomised studies get $\mathcal N(0, \kappa^2 s_i^2)$. The current default. |
| `fixed_zero` | `fixed_zero` | `zero` | $\gamma = 0$ for post-only studies (naive comparator). |

What each block asks:

- **X1–X3 (transport).** Unrandomised DiD studies whose treated arms all start high, alongside genuinely randomised trials. The `shared` arm should be biased by roughly the DiD imbalance, and that bias should *not* shrink as DiD evidence sharpens — it should sharpen with it.
- **X4–X6 (direction).** Same imbalance magnitudes, sign flipped study by study. `shared` should now be roughly unbiased in the mean while $\tau_\gamma$ absorbs the full magnitude. This isolates whether $\mathcal N(0, \tau_\gamma)$ is the *right* amount of conservatism or merely a safe one.
- **X7–X9 (mislabelling).** Prices the cost of the `randomisation` claim being wrong, and asks whether a larger $\kappa$ buys the robustness back.
- **X10–X12 (hard-zero trap).** $\kappa = 0$ pins $\gamma$ at zero, so a real imbalance can no longer be differenced away. Compares against shrinking toward zero ($\kappa > 0$) and leaving $\gamma$ free.
- **X13–X15 ($\kappa$ identification).** $\kappa$ is anchored only by randomised studies carrying pre-treatment data. X13 has none, so it sweeps fixed values instead — the spread across that sweep *is* the sensitivity band that justifies the default.
- **X16–X18 (cluster randomisation).** The design effect is applied to the baseline contrast only; the post-period likelihood still uses $\sigma^2/n$, so residual over-precision is expected. Measuring its size is the point.
- **X19 (aliasing).** For post-only studies only $\delta_{\mathrm{RCT}} + \mu_\gamma$ is identified.
- **X20–X22 (normalisation).** $\gamma \equiv 0$ throughout, so $\tau_\gamma$ should concentrate at zero. Under normalisation the data are divided by the *observed* control baseline, which the likelihood then treats as exact — dropping that divisor's sampling variance from the baseline contrast. If it matters, $\tau_\gamma$ should sit above zero under normalisation and shrink with $n$. **$\tau_\gamma$ is exactly the quantity the two-population model transports to post-only studies, so this is a prerequisite for trusting the rest of X.**
- **X24–X27 (pre-post scale mismatch).** A PP study has no control arm, so it normalises by its own treatment-arm pre mean while DiD and RCT normalise by the control baseline. When the treated group is selected those denominators differ by $(1 + \gamma/b)$, yet all designs pool into one $\mu_\theta$. $\gamma$ is swept with the PP share held at half, so X24 ($\gamma = 0$) is a built-in null, and the unnormalised arm is a second control — pooling on the absolute scale involves no divisor, so its slope across $\gamma$ should be flat. **Deliberately not registered in `scenario_expectations()`:** this is a real limitation rather than a naive comparator, so it belongs under "Genuine issues".
- **X23 (individual data).** For summary studies $s_i$ is computed in R from reported SDs and passed as data; for individual studies the observation SDs are *parameters*, so $s_i$ is built inside Stan and its uncertainty propagated. Without this the branch ships untested.

Arms marked misspecified-by-design in `scenario_expectations()` are expected to deviate — for X1–X3 in particular, the `shared` arm's bias *is* the result.

```{r x-scenarios}
scenario_summary_table("X") |>
  kable(caption = "Category X: scenario definitions")
```

```{r x-diagnostics}
#| fig-height: 8
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "X")
```

```{r x-table}
results_table(all_agg, scenario_lookup_tbl, "X")
```

## Convergence Diagnostics

```{r convergence}
plot_convergence(all_agg)
```

### Sampler divergences

The HMC sampler reports a divergent transition when its leapfrog
integrator overshoots a sharp curve in the posterior. A few divergences
in a long run are usually tolerable; many divergences flag a posterior
geometry the sampler can't navigate (typically funnels, hard
constraints near the posterior bulk, or wide-uniform-prior pathologies).

```{r divergences}
#| fig-height: 8
plot_divergences(all_agg)
```

## Full Results Table

```{r full-table}
all_agg |>
  arrange(scenario_id, model_label, parameter) |>
  kable(digits = 3, caption = "All aggregated results")
```
