---
title: "metadid Simulation Validation Report"
author: "metadid-sims pipeline"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
execute:
  echo: true
  warning: false
---

```{r setup}
library(targets)
library(dplyr)
library(ggplot2)
library(knitr)

tar_source("R/scenarios.R")
tar_source("R/plots.R")

all_agg <- tar_read(all_agg)
all_rep <- tar_read(all_rep)
```

## Executive Summary

```{r exec-summary}
# Flag scenarios with empirical coverage below 80% for the treatment effect
poor_coverage <- all_agg |>
  filter(parameter == "treatment_effect_mean", empirical_coverage < 0.80)

convergence_issues <- all_agg |>
  filter(parameter == "treatment_effect_mean", pct_rhat_bad > 0.10)
```

**`r nrow(all_agg |> distinct(scenario_id))`** scenarios evaluated across
**`r max(all_agg$n_reps, na.rm = TRUE)`** replications each.

`r if (nrow(poor_coverage) == 0) "All scenarios achieved ≥80% empirical coverage for the treatment effect mean." else paste0(nrow(poor_coverage), " scenario(s) had coverage below 80% for the treatment effect mean.")`

`r if (nrow(convergence_issues) == 0) "No convergence issues detected (all scenarios had <10% of fits with Rhat > 1.05)." else paste0(nrow(convergence_issues), " scenario(s) had >10% of fits with convergence warnings.")`

```{r poor-coverage-table}
#| eval: !expr nrow(poor_coverage) > 0
kable(
  poor_coverage |>
    left_join(
      tibble(
        scenario_id = names(SCENARIO_CONFIGS),
        description = purrr::map_chr(SCENARIO_CONFIGS, "description")
      ),
      by = "scenario_id"
    ) |>
    select(scenario_id, description, model_label, empirical_coverage, mean_bias, rmse),
  digits = 3,
  caption = "Scenarios with <80% coverage for treatment_effect_mean"
)
```

## A. Calibration Studies

```{r a-coverage}
plot_coverage(all_agg, category = "A")
```

```{r a-bias}
plot_bias(all_rep, category = "A")
```

```{r a-table}
all_agg |>
  filter(grepl("^A", scenario_id)) |>
  select(scenario_id, model_label, parameter,
         empirical_coverage, mean_bias, rmse, mean_ci_width) |>
  kable(digits = 3, caption = "Category A: Calibration results")
```

## F. Large-N Bias Probes

```{r f-coverage}
plot_coverage(all_agg, category = "F")
```

```{r f-bias}
plot_bias(all_rep, category = "F")
```

```{r f-table}
all_agg |>
  filter(grepl("^F", scenario_id)) |>
  select(scenario_id, model_label, parameter,
         empirical_coverage, mean_bias, rmse, mean_ci_width) |>
  kable(digits = 3, caption = "Category F: Large-N bias probe results")
```

## B. Comparative Studies

```{r b-comparison}
plot_comparison(all_rep |> filter(grepl("^B", scenario_id)))
```

```{r b-table}
all_agg |>
  filter(grepl("^B", scenario_id)) |>
  select(scenario_id, model_label, parameter,
         empirical_coverage, mean_bias, rmse) |>
  kable(digits = 3, caption = "Category B: Model comparison results")
```

## C. Outlier and Heavy-Tailed Studies

```{r c-comparison}
plot_comparison(all_rep |> filter(grepl("^C", scenario_id)))
```

```{r c-table}
all_agg |>
  filter(grepl("^C", scenario_id)) |>
  select(scenario_id, model_label, parameter,
         empirical_coverage, mean_bias, rmse) |>
  kable(digits = 3, caption = "Category C: Outlier/heavy-tail results")
```

## D. Assumption Violation Studies

```{r d-coverage}
plot_coverage(all_agg, category = "D")
```

```{r d-bias}
plot_bias(all_rep, category = "D")
```

```{r d-table}
all_agg |>
  filter(grepl("^D", scenario_id)) |>
  select(scenario_id, model_label, parameter,
         empirical_coverage, mean_bias, rmse) |>
  kable(digits = 3, caption = "Category D: Assumption violation results")
```

## E. Edge Cases

```{r e-coverage}
plot_coverage(all_agg, category = "E")
```

```{r e-table}
all_agg |>
  filter(grepl("^E", scenario_id)) |>
  select(scenario_id, model_label, parameter,
         empirical_coverage, mean_bias, rmse) |>
  kable(digits = 3, caption = "Category E: Edge case results")
```

## Convergence Diagnostics

```{r convergence}
plot_convergence(all_agg)
```

## Full Results Table

```{r full-table}
all_agg |>
  arrange(scenario_id, model_label, parameter) |>
  kable(digits = 3, caption = "All aggregated results")
```
