---
title: "metadid Simulation Validation Report"
author: "metadid-sims pipeline"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
    page-layout: full
execute:
  echo: true
  warning: false

knitr:
  opts_chunk:
    fig.width: 16
    fig.height: 7
    out.width: "100%"
---

```{r setup}
library(targets)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)

tar_source("R/scenarios.R")
tar_source("R/plots.R")

all_agg <- tar_read(all_agg)
all_rep <- tar_read(all_rep)
scenario_lookup_tbl <- scenario_lookup()
```

## Executive Summary

```{r exec-summary}
# Flag scenarios with empirical coverage below 80% for the treatment effect
poor_coverage <- all_agg |>
  filter(parameter == "treatment_effect_mean", empirical_coverage < 0.80)

convergence_issues <- all_agg |>
  filter(parameter == "treatment_effect_mean", pct_rhat_bad > 0.10)
```

**`r nrow(all_agg |> distinct(scenario_id))`** scenarios evaluated across
**`r max(all_agg$n_reps, na.rm = TRUE)`** replications each.

`r if (nrow(poor_coverage) == 0) "All scenarios achieved ≥80% empirical coverage for the treatment effect mean." else paste0(nrow(poor_coverage), " scenario(s) had coverage below 80% for the treatment effect mean.")`

`r if (nrow(convergence_issues) == 0) "No convergence issues detected (all scenarios had <10% of fits with Rhat > 1.05)." else paste0(nrow(convergence_issues), " scenario(s) had >10% of fits with convergence warnings.")`

```{r poor-coverage-table}
#| eval: !expr nrow(poor_coverage) > 0
kable(
  poor_coverage |>
    left_join(
      tibble(
        scenario_id = names(SCENARIO_CONFIGS),
        description = purrr::map_chr(SCENARIO_CONFIGS, "description")
      ),
      by = "scenario_id"
    ) |>
    select(scenario_id, description, model_label, empirical_coverage, mean_bias, rmse),
  digits = 3,
  caption = "Scenarios with <80% coverage for treatment_effect_mean"
)
```

## A. Calibration Studies

These scenarios test the model's core calibration properties — coverage, bias, and RMSE — under a range of study counts, design mixes, and parameterisations. All use the full `meta_did` model unless noted. The baseline scenario (A1) uses 20 DiD studies with default parameters; subsequent scenarios probe how performance changes with sample size (A2–A4), mixed designs (A5–A6), covariates (A7), correlated effects (A8), unnormalised fitting (A9–A10), individual-level data (A11–A12), rho estimation strategies (A13), and sign/direction variants (A14–A18).

```{r a-scenarios}
scenario_summary_table("A") |>
  kable(caption = "Category A: scenario definitions")
```

```{r a-diagnostics}
#| fig-height: 9
#| column: screen
#| out-width: "100%"
A_combined    <- tar_read(A_combined)
A_agg_combined <- tar_read(A_agg_combined)
plot_diagnostics(A_combined, A_agg_combined, scenario_lookup_tbl, "A")
```

```{r a-table}
results_table(all_agg, scenario_lookup_tbl, "A")
```

## F. Large-N Bias Probes

These scenarios use large numbers of studies (200 total) to narrow posteriors and expose any systematic bias that would be obscured in smaller samples. They vary the design mix, normalisation strategy, and model specification, allowing a stringent test of whether the model is asymptotically unbiased.

```{r f-scenarios}
scenario_summary_table("F") |>
  kable(caption = "Category F: scenario definitions")
```

```{r f-diagnostics}
#| fig-height: 7
#| column: screen
#| out-width: "100%"
F_combined     <- tar_read(F_combined)
F_agg_combined <- tar_read(F_agg_combined)
plot_diagnostics(F_combined, F_agg_combined, scenario_lookup_tbl, "F")
```

```{r f-table}
results_table(all_agg, scenario_lookup_tbl, "F")
```

## B. Comparative Studies

These scenarios pit two model configurations against each other on the same simulated data, asking which recovers the treatment effect better under specific conditions. Each scenario has a clear hypothesis about which model should win (or whether they should tie).

```{r b-scenarios}
scenario_summary_table("B") |>
  kable(caption = "Category B: scenario definitions")
```

```{r b-diagnostics}
#| fig-height: 7
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "B")
```

```{r b-table}
results_table(all_agg, scenario_lookup_tbl, "B")
```

## C. Outlier and Heavy-Tailed Studies

These scenarios test robustness to departures from Gaussian study-level heterogeneity. The contaminating distributions are chosen to be detectable but not catastrophic, so that a robust model (Student-t heterogeneity) should outperform the normal model without being trivially obvious.

```{r c-scenarios}
scenario_summary_table("C") |>
  kable(caption = "Category C: scenario definitions")
```

```{r c-diagnostics}
#| fig-height: 8
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "C")
```

```{r c-table}
results_table(all_agg, scenario_lookup_tbl, "C")
```

## D. Assumption Violation Studies

These scenarios introduce deliberate mismatches between the DGP and the fitted model to quantify the cost of specific assumption violations — unmodelled design effects, heterogeneous within-study variance, misspecified rho, effect-size/sample-size correlation, and ignored RCT baseline imbalance.

```{r d-scenarios}
scenario_summary_table("D") |>
  kable(caption = "Category D: scenario definitions")
```

```{r d-diagnostics}
#| fig-height: 6
#| column: screen
#| out-width: "100%"
D_combined     <- tar_read(D_combined)
D_agg_combined <- tar_read(D_agg_combined)
plot_diagnostics(D_combined, D_agg_combined, scenario_lookup_tbl, "D")
```

```{r d-table}
results_table(all_agg, scenario_lookup_tbl, "D")
```

## E. Edge Cases

These scenarios probe boundary conditions: extreme pre-post correlations, zero heterogeneity, very large heterogeneity, unbalanced arms, single studies per design, and missing rho values. They test whether the model remains stable and well-calibrated when inputs approach the limits of the assumed data-generating regime.

```{r e-scenarios}
scenario_summary_table("E") |>
  kable(caption = "Category E: scenario definitions")
```

```{r e-diagnostics}
#| fig-height: 6
#| column: screen
#| out-width: "100%"
E_combined     <- tar_read(E_combined)
E_agg_combined <- tar_read(E_agg_combined)
plot_diagnostics(E_combined, E_agg_combined, scenario_lookup_tbl, "E", exclude = "E5")
```

```{r e-table}
results_table(all_agg, scenario_lookup_tbl, "E")
```

## H. Time Trend Distributional Misspecification

The full model assumes study-level time trends are exchangeable draws from a normal distribution. These scenarios test what happens when that assumption is violated at the population level, using a 15 DiD + 15 PP design so that PP studies — which rely on trend estimation for effect identification — are directly exposed to the misspecification. Each scenario compares three models on the same simulated data: the **full normal** model (`meta_did`), the **full robust** model (`meta_did` with Student-t heterogeneity), and the **naive** model (`meta_did_general` with `time_trend = "fixed_zero"`).

H1 generates trends from a negated log-normal (mean = −0.04, log-SD = 1), producing strong right-skewness: most studies have trends close to zero with occasional large negative trends. H2 uses a 50/50 mixture of two normal components centred at −0.01 and −0.10 — a bimodal distribution the normal model cannot represent regardless of sample size. H3 keeps the normal DGP but inflates `sigma_trend` to 0.10 (five times the default), testing the regime where trend uncertainty dominates effect uncertainty. H4 introduces a design-confounded trend: DiD studies are small (`n ~ U(30, 70)`) and have more negative trends, while PP studies are large (`n ~ U(150, 250)`) and have trends closer to zero; because the model pools trends across designs, it over-adjusts PP outcomes.

```{r h-scenarios}
scenario_summary_table("H") |>
  kable(caption = "Category H: scenario definitions")
```

```{r h-diagnostics}
#| fig-height: 7
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "H")
```

```{r h-table}
results_table(all_agg, scenario_lookup_tbl, "H")
```

## G. Bias Source Investigation

Category G probes the two systematic bias sources identified in the large-N results (Category F). It comprises two sub-investigations.

**G1–G5: Prior sensitivity (unnormalised model).** The unnormalised model showed a small positive bias (~0.6–0.9%) for a negative true effect, consistent with hierarchical shrinkage toward zero from the default `N(0, 10)` prior on `treatment_effect_mean`. G1–G5 vary the prior standard deviation from 1 to 100 (G3 replicates the default) to test whether the bias decreases monotonically as the prior becomes more diffuse.

**G6–G9: Jensen's inequality (normalised model).** The normalised model showed ~1% systematic downward bias at 200 studies, consistent with Jensen's inequality on the ratio estimator: because the model fits E[θ/baseline] rather than E[θ]/E[baseline], the convexity of the reciprocal function induces a negative shift proportional to between-study baseline variance. G6–G8 vary between-study baseline variation (`baseline_sd` = 0, 0.01, 0.10) to confirm that the bias tracks the Jensen term; G9 uses a very large control group (`n_control = 5000`) to test whether high within-study precision alone eliminates the bias.

```{r g-scenarios}
scenario_summary_table("G") |>
  kable(caption = "Category G: scenario definitions")
```

```{r g-diagnostics}
#| fig-height: 9
#| column: screen
#| out-width: "100%"
plot_diagnostics(all_rep, all_agg, scenario_lookup_tbl, "G")
```

```{r g-table}
results_table(all_agg, scenario_lookup_tbl, "G")
```

## Convergence Diagnostics

```{r convergence}
plot_convergence(all_agg)
```

## Full Results Table

```{r full-table}
all_agg |>
  arrange(scenario_id, model_label, parameter) |>
  kable(digits = 3, caption = "All aggregated results")
```
